MCMC convergence diagnostics

### 2023-12-04 src/naomi-simple_mcmc/mcmc-convergence.Rmd
Update NUTS diagnostics for thesis

```{r}
rhats <- bayesplot::rhat(mcmc)

bayesplot::mcmc_rhat_data(rhats) %>%
  ggplot(aes(x = value, y = parameter, col = description)) +
  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7) +
  scale_color_manual(values = "#E69F00") +
  geom_vline(xintercept = 1.05, linetype = "dashed", col = "grey40") +
  labs(x = "Potential scale reduction factor", y = "NUTS parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank()
  )

@@

(max_rhat <- max(rhats))
```

# ESS ratio

@@

bayesplot::mcmc_neff_data(ratios) %>%
  ggplot(mapping = aes(x = value, y = parameter, color = description)) +
  geom_segment(aes(yend = parameter, xend = -Inf), na.rm = TRUE, alpha = 0.7) +
  scale_color_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40") +
  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40") +
  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40") +
  labs(x = "ESS ratio", y = "NUTS parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank()
  )

ggsave("ratio.png", h = 3, w = 6.25)

(average_ess_ratio <- mean(ratios))
```

@@

```{r}
#' I think that this $summary should be all of the chains grouped together
mcmc_summary <- summary(mcmc)$summary

data.frame(mcmc_summary) %>%
  tibble::rownames_to_column("param") %>%
  ggplot(aes(x = n_eff)) +
    geom_histogram(alpha = 0.8) +
    labs(x = "ESS", y = "Count")

ggsave("ess.png", h = 3, w = 6.25)

(ess_min <- min(mcmc_summary[, "n_eff"]))
(ess_lower <- quantile(mcmc_summary[, "n_eff"], 0.025))
(ess_median <- quantile(mcmc_summary[, "n_eff"], 0.50))
(ess_upper <- quantile(mcmc_summary[, "n_eff"], 0.975))
(ess_max <- max(mcmc_summary[, "n_eff"]))
```

@@

```{r}
out <- list(
  "max_rhat" = max_rhat,
  "average_ess_ratio" = average_ess_ratio, 
  "ess_min" = ess_min,
  "ess_lower" = ess_lower,
  "ess_median" = ess_median,
  "ess_upper" = ess_upper,
  "ess_max" = ess_max
)

@@

```{r}
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("logit")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))
```

@@

ggsave("worst-trace.png", plot, h = 4, w = 6.25)
```

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_as", "logit_phi_alpha_as"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")
(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a"), diag_fun = "hist", off_diag_fun = "hex"))

ggsave("rho_a.png", plot, h = 4, w = 6.25)

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_x", "logit_phi_rho_x"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_xs", "logit_phi_rho_xs"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")
(plot <- bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x"), diag_fun = "hist", off_diag_fun = "hex"))

ggsave("alpha_x.png", plot, h = 4, w = 6.25)

>>>>

```{r}
rhats <- bayesplot::rhat(mcmc)
rhats <- bayesplot:::drop_NAs_and_warn(bayesplot:::new_rhat(rhats))
rhats_df <- bayesplot:::diagnostic_data_frame(rhats)

ggplot(rhats_df, aes(x = value, y = parameter)) +
  geom_segment(aes(yend = parameter, xend = ifelse(min(value) < 1, 1, -Inf)), na.rm = TRUE, alpha = 0.7, col = "#56B4E9") +
  lims(x = c(NA, 1.03)) +
  labs(x = "Potential scale reduction factor", y = "Parameter", col = "") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

@@

#' I think that this $summary should be all of the chains grouped together
mcmc_summary <- summary(mcmc)$summary

(rhat_min <- min(mcmc_summary[, "Rhat"]))
(rhat_lower <- quantile(mcmc_summary[, "Rhat"], 0.025))
(rhat_median <- quantile(mcmc_summary[, "Rhat"], 0.50))
(rhat_upper <- quantile(mcmc_summary[, "Rhat"], 0.975))
(rhat_max <- max(mcmc_summary[, "Rhat"]))
(rhat_mean <- mean(mcmc_summary[, "Rhat"]))
```

@@

ess_fig <- bayesplot::mcmc_neff_data(ratios) %>%
  mutate(
    description_new = case_when(
      value < 0.1 ~ "< 0.1",
      value < 0.5 ~ "< 0.5",
      value > 0.5 ~ "> 0.5"
    )
  ) %>%
  ggplot(mapping = aes(x = value, y = parameter, color = description_new)) +
  geom_segment(aes(yend = parameter, xend = -Inf), na.rm = TRUE, alpha = 0.7) +
  scale_color_manual(values = c("#E69F00", "#009E73", "#56B4E9")) +
  geom_vline(xintercept = 0.1, linetype = "dashed", col = "grey40", alpha = 0.5) +
  geom_vline(xintercept = 0.5, linetype = "dashed", col = "grey40", alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed", col = "grey40", alpha = 0.5) +
  labs(x = "ESS ratio", y = "Parameter", col = "", tag = "A") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )

```

@@

```{r}
ess_histogram <- data.frame(mcmc_summary) %>%
  tibble::rownames_to_column("param") %>%
  ggplot(aes(x = n_eff)) +
    geom_histogram(alpha = 0.8) +
    labs(x = "ESS", y = "Count", tag = "B")

ess_fig / ess_histogram

ggsave("ess.png", h = 5, w = 6.25)

(ess_min <- min(mcmc_summary[, "n_eff"]))
(ess_lower <- quantile(mcmc_summary[, "n_eff"], 0.025))
(ess_median <- quantile(mcmc_summary[, "n_eff"], 0.50))
(ess_upper <- quantile(mcmc_summary[, "n_eff"], 0.975))
(ess_max <- max(mcmc_summary[, "n_eff"]))
(ess_mean <- mean(mcmc_summary[, "n_eff"]))
```

@@

```{r}
out <- list(
  "rhat_min" = rhat_min,
  "rhat_lower" = rhat_lower,
  "rhat_median" = rhat_median,
  "rhat_upper" = rhat_upper,
  "rhat_max" = rhat_max,
  "rhat_mean" = rhat_mean,
  "ess_min" = ess_min,
  "ess_lower" = ess_lower,
  "ess_median" = ess_median,
  "ess_upper" = ess_upper,
  "ess_max" = ess_max,
  "ess_mean" = ess_mean
)

@@

```{r}
cbpalette <- c("#56B4E9", "#009E73", "#E69F00", "#F0E442", "#E69F00", "#E69F00")
bayesplot::color_scheme_set(rev(cbpalette))

bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("beta")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("logit")))
bayesplot::mcmc_trace(mcmc, pars = vars(starts_with("log_sigma")))
```

@@

ggsave("worst-trace.png", plot, h = 3, w = 6.25)
```

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_a", "logit_phi_alpha_a"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_as", "logit_phi_alpha_as"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_as", "logit_phi_rho_as"), diag_fun = "hist", off_diag_fun = "hex")

pairs_df <- extract(mcmc, pars = c("log_sigma_rho_a", "logit_phi_rho_a")) %>%
  as.data.frame()

fig_top_left <- ggplot(pairs_df, aes(x = log_sigma_rho_a)) +
  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
  labs(x = "", y = "", tag = "A") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_top_right <- ggplot(pairs_df, aes(x = logit_phi_rho_a, y = log_sigma_rho_a)) +
  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
  labs(x = "", y = "", tag = "B") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_bot_left <- ggplot(pairs_df, aes(x = log_sigma_rho_a, y = logit_phi_rho_a)) +
  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
  labs(x = "log_sigma_rho_a", y = "", tag = "C") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_bot_right <- ggplot(pairs_df, aes(x = logit_phi_rho_a)) +
  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
  labs(x = "logit_phi_rho_a", y = "", tag = "D") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

(fig_top_left + fig_top_right) / (fig_bot_left + fig_bot_right)

ggsave("rho_a.png", h = 4, w = 6.25)

@@

```{r}
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_x", "logit_phi_rho_x"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_rho_xs", "logit_phi_rho_xs"), diag_fun = "hist", off_diag_fun = "hex")
bayesplot::mcmc_pairs(mcmc, pars = c("log_sigma_alpha_xs", "logit_phi_alpha_xs"), diag_fun = "hist", off_diag_fun = "hex")

pairs_df <- extract(mcmc, pars = c("log_sigma_alpha_x", "logit_phi_alpha_x")) %>%
  as.data.frame()

fig_top_left <- ggplot(pairs_df, aes(x = log_sigma_alpha_x)) +
  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
  labs(x = "", y = "", tag = "A") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_top_right <- ggplot(pairs_df, aes(x = logit_phi_alpha_x, y = log_sigma_alpha_x)) +
  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
  labs(x = "", y = "", tag = "B") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_bot_left <- ggplot(pairs_df, aes(x = log_sigma_alpha_x, y = logit_phi_alpha_x)) +
  geom_point(shape = 1, alpha = 0.5, col = "#009E73") +
  labs(x = "log_sigma_alpha_x", y = "", tag = "C") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

fig_bot_right <- ggplot(pairs_df, aes(x = logit_phi_alpha_x)) +
  geom_histogram(fill = "#56B4E9", col = "#1a88c6") +
  labs(x = "logit_phi_alpha_x", y = "", tag = "D") + 
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

(fig_top_left + fig_top_right) / (fig_bot_left + fig_bot_right)

ggsave("alpha_x.png", h = 4, w = 6.25)
